---
title: "Prerequisite assignment"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**This assignment is due Nov 29th before midnight**  

Please read the instructions and watch the video (~20 min) provided in the [R package installation and Rstudio overview](https://rnabioco.github.io/bmsc-7810-pbda/posts/2023-11-06-install-r).

This assignment is designed to check that your R and Rstudio installation is set up correctly. If this assignment is taking more than 1 hour of your time please contact the instructors for assistance (rbi.fellows@cuanschutz.edu).

Please perform the following:

1) Change the title of this Rmarkdown to include your name e.g. (`Kent Riemondy Prerequisite assignment`)

2) Add a R code chunk to the document and load the `tidyverse` package, using the `library()` function.

3) Knit this R Markdown document to generate HTML output (click the `Knit` needle and yarn icon).

4) Submit the html document to Canvas. 


Keep the following code chunk in your R Markdown, as it will demonstrate which packages you've installed. 

```{r}
sessionInfo()
```

