---
title: "Class 11: Programming in R (part 2)"
author:
  - name: Ryan Sheridan and Kent Riemondy
date: 2022-12-05
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "left"
)
```

*The Rmarkdown for this class is [on github]( https://github.com/rnabioco/bmsc-7810-pbda/blob/main/_posts/2022-12-05-class-11-programming-in-r-pt2/programming-in-r-pt2.Rmd)*


```{r}
# conditionally download all of the files used in rmarkdown from github 
source("https://github.com/rnabioco/bmsc-7810-pbda/raw/main/_posts/2022-12-05-class-10-programming-in-r-part-1/download-files.R")
```


<br>

## Introduction

As discussed in the previous class, you should try to limit duplication in your code. One way to do this is by writing functions, another way is through iteration. Reducing code duplication has several benefits:

* Your code easier to read
* You reduce the chance of mistakes from repeated copying and pasting
* It is easier to adapt your code for different requirements

```{r}
library(tidyverse)
library(patchwork)
```

<br>

<br>


## `for` loops
`for` loops allow you to run the same code block repeatedly without copying and pasting. 


```{r}
for(i in 1:4){
  message("printing ", i, " is fun")
}

values <- c("A", "B", "C", "D")
for(val in values){
  message(val)
}
```


When iterating over a vector, usually it is most useful to iterate over the index of each element (aka the position in the vector), rather than the values themselves. 

```{r}
for(i in 1:length(values)){
  val <- values[i]
  message("index = ", i, " value = ", val)
}
```

We will generally want to store the output generated in the `for` loop. A common paradigm is
to preallocate a place to store the output. This is a faster approach than growing the output at each iteration (for more detail see [this R-bloggers post](https://www.r-bloggers.com/2018/08/growing-objects-and-loop-memory-pre-allocation/)). 

We can generate vectors (and lists) of a given type and length using the `vector()` function. 

```{r}
n <- length(values)

# make an empty vector of same length as values (4)
output <- vector(mode = "character", length = n)
output

for(i in 1:n){ 
  # get value at position i
  val <- values[i]
  # assign value to output character vector at position i
  output[i] <- tolower(val)
}
output
```

```{r}
# It's helpful to think about what happens during each cycle of the loop
output[1] <- tolower("A") # i = 1
output[2] <- tolower("B") # i = 2
output[3] <- tolower("C") # i = 3
output[4] <- tolower("D") # i = 4
```

<br> 

<br> 

Lets use `rnorm()` to create a list of 5 vectors with different values for 'mean'

```{r}
# One way to do this is by copying and pasting
vec_in <- c(1, 50, 20, 5, 70)                # input

out <- vector("list", length(vec_in))        # output

out[[1]] <- rnorm(n = 10, mean = vec_in[1])    
out[[2]] <- rnorm(n = 10, mean = vec_in[2])
out[[3]] <- rnorm(n = 10, mean = vec_in[3])
out[[4]] <- rnorm(n = 10, mean = vec_in[4])
out[[5]] <- rnorm(n = 10, mean = vec_in[5])
out
```

```{r}
# Use a for loop to reduce code duplication
vec_in <- c(1, 50, 20, 5, 70)                  # input

out <- vector("list", length(vec_in))          # output

for (i in 1:length(vec_in)) {                  # sequence
  
  out[[i]] <- rnorm(n = 10, mean = vec_in[i])  # body
  
}
```

<br>

*Write a `for` loop that uses `rnorm()` to create 3 vectors of different lengths. Store the vectors in a list. Use mean = 0 and sd = 1 (the default). *  

<details>
  <summary>Show answer</summary>
  
```{r}
vec_in <- c(5, 10, 2)                  # input
n <- length(vec_in)
out <- vector("list", n)  # output

for (i in 1:length(vec_in)) {          # sequence
  out[[i]] <- rnorm(n = vec_in[i])
}

out
```

</details>

<br>

So far we have used `1:length(x)` to specify the sequence to iterate over. A better alternative is using `seq_along(x)` instead of `1:length(x)` . This guards against errors when an empty vector is passed to `1:length(x)`.

```{r}
# seq_along() mimics 1:length() for non-empty vectors
vec_in <- c(5, 10, 2)

1:length(vec_in)

seq_along(vec_in)

# seq_along() correctly handles empty vectors
emp_vec <- vector("numeric", 0)

1:length(emp_vec)

seq_along(emp_vec)
```

<br>

<br>

## Using the Brauer data

Using the Brauer gene expression data lets create a figure showing the growth rate vs expression for four genes

```{r}
brauer_gene_exp <- read_csv("data/brauer_gene_exp.csv.gz")
```

```{r}
# This is the function we wrote in class-10
plot_expr <- function(input, sys_name, ...) {
  
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(...) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme(legend.position = "none")
}

```

Lets try this with the copy-and-paste method, storing the plots in a list.

```{r, fig.width = 10, fig.height = 8}
vec_in <- c("YDL104C", "YLR115W", "YMR183C", "YML017W")       # input

out <- vector("list", length(vec_in))                         # output 

out[[1]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[1])  
out[[2]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[2])
out[[3]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[3])
out[[4]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[4])

wrap_plots(out)
```

<br>

*Re-write the code from above using a `for` loop to generate our figure*  

<details>
  <summary>Show answer</summary>
  
```{r, eval = F}
vec_in <- c("YDL104C", "YLR115W", "YMR183C", "YML017W")  # input

out <- vector("list", length(vec_in))                    # output

for (i in seq_along(vec_in)) {                                      
  out[[i]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[i])      
}
out
wrap_plots(out)
```

</details>

<br>

<br>

## A note on vectorization

*In general you should try to use a vectorized function/approach before using iteration.* Vectorized approaches will be faster and require less code to run. If you are working with a vector or matrix, then there is likely a vectorized operation that can be used. 

There are however a few common places that iteration is used:  
  - To process multiple datasets/data.frames. (e.g. apply a function to a list of data.frames)     
  - plotting (e.g. make many plots with a varying input parameter or aesthetic)  
  - perform a custom operation that does not have a vectorized approach 


## Introduction to the `apply` family of functions.

`for` loops are a powerful tool to reduce code duplication, however your code can be  simplified using the `lapply` function and related apply functions in base R. These functions essentially run `for (i in seq_along(x))` behind the scenes so you don't have to explicitly type this.

There is a function for each type of output:

- `lapply()`  iterate over a vector, applying a function, returning a list  
- `sapply()`  iterate over a vector, applying a function, coercing the output to a vector  
- `apply()`  iterate over a row, column, or all elements of a matrix  
- `mapply()` iterate over each element of multiple supplied vectors   

The `lapply` function requires two inputs: `lapply(X, FUN, ...)`

- `X` is a list or atomic vector
- `FUN` is a function
- ... additional arguments to FUN  


```{r}
# We previously used a for loop to create vectors with different values for mean
vals <- c(1, 50, 20, 5, 70)                  # input

out <- vector("list", length(vals))          # output

for (i in seq_along(vals)) {                 # sequence
  
  out[[i]] <- rnorm(n = 10, mean = vals[i])  # body
  
}

# Using lapply() we can further simplify this code
# x indicates where each element of the vector should be inserted
# this is an example of an anonymous function
out <- lapply(vals, function(x) rnorm(n = 10, mean = x))

# we can also define the function first
rnorm_custom <- function(x){
  rnorm(n = 10, mean = x)
}

out <- lapply(vals, rnorm_custom)

# You can use brackets to include a multi-line code block
out <- lapply(vals, function(x) {
  
  rnorm(n = 10, mean = x)
  
})

# Each element of the vector is passed to the first available argument
out <- lapply(vals, rnorm, n = 10)
#out <- lapply(vals, rnorm)
```

<br>

*Use `rnorm()` and `lapply()` to create 3 vectors of different lengths*

<details>
  <summary>Show answer</summary>
  
```{r}

out <- lapply(c(10, 1, 4), rnorm)
out <- lapply(c(10, 1, 4), function(x) rnorm(x))
rnorm2 <- function(x){
  rnorm(n = x)
}
out <- lapply(c(10, 1, 4), function(x) rnorm2(x))
out <- lapply(c(10, 1, 4), rnorm2)
out
```

</details>

<br>

*Re-write the code from above using `lapply()` to generate our growth rate figure*

<details>
  <summary>Show answer</summary>
```{r, eval = F, fig.width = 10, fig.height = 8}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")

plot_expr(brauer_gene_exp, sys_name = vec_in[i])   

expr_plots <- lapply(genes, plot_expr, input = brauer_gene_exp)

expr_plots <- lapply(genes, function(gene){
  plot_expr(brauer_gene_exp, sys_name = gene)
})

expr_plots
wrap_plots(expr_plots)
```

</details>

<br>

<br>

Note that we can iterate over lists in addition to vectors. A common operation might be to read in multiple files and perform some operation

```{r, eval = FALSE}
# get paths to files in "data" directory (dir() is an alias for list.files())
file_names <- dir("data", full.names = TRUE)

# read each file into R and store in a list
lst_of_dfs <- lapply(file_names, read_csv)

# get nrow of each file
lapply(lst_of_dfs, nrow)

# select 5 random rows
lapply(lst_of_dfs, slice_sample, n = 5)

# check if any NAs are present
lapply(lst_of_dfs, function(x){
  sum(is.na(x)) > 0
})
```

Note that a `data.frame` is a list in R, such that each column is one element of a list (e.g. see output of `typeof(mtcars)`). So if we use `lapply()` on a data.frame it will iterate over each column.

```{r}
lapply(mtcars, mean)
lapply(mtcars, class)
```


`sapply()` will coerce the output to be a vector rather than a list.

```{r}
sapply(mtcars, mean)
lapply(mtcars, mean) %>% unlist()
```

## iterating over matrices

The `apply()` function allows you to iterates over a row, column, or all elements of a matrix.

```{r}
mat <- matrix(rnorm(100),
              nrow = 10,
              dimnames = list(paste0("row_", letters[1:10]),
                              paste0("column_", letters[11:20])))
mat
```

The arguments are:  
`apply(X, MARGIN, FUN, ...)`  

X = matrix    
MARGIN = specify how to iterate (1 = by row, 2 = by column, c(1,2) = all values)  
FUN = function to apply  

```{r, eval = FALSE}
# get max value in each row
apply(mat, 1, max)

# get max value in each column
apply(mat, 2, max)

# multiply each value by 1e6  
apply(mat, c(1, 2), function(x) x * 1e6)
```

Alternatively you can use for loops to iterate over matrices by using indexing to select/replace elements

```{r, eval = FALSE}
nr <- nrow(mat)
nc <- ncol(mat)

# iterate by column, seq_len makes range 1:nc
output <- vector("numeric", nc)
for(i in seq_len(nc)){
  col <- mat[, i]
  output[i] <- sum(col + 5)
}
```

```{r, eval = FALSE}
# iterate by row, seq_len makes range 1:nr 
output <- vector("numeric", nr)
for(i in seq_len(nr)){
  row <- mat[i, ]
  output[i] <-  sum(row + 5)
}
```

```{r, eval = FALSE}
# iterate by row, and column using nested for loops

# make an output matrix, filled with 0s, with same names as mat
output <- matrix(0, nrow = nr, ncol = nc, dimnames = dimnames(mat))

for(i in seq_len(nr)){
  for(j in seq_len(nc)){
    val <- mat[i, j]
    output[i, j] <- val * 1e6
  }
}
```

**Note that many operations on matrices are vectorized already and you often don't need to use apply or for loops**

```{r, eval = FALSE}
mat * 1e6
```

Use the `matrixStats` package for common row-wise or column-wise operations before using apply. 

```{r}
# in base R
rowMeans(mat)

# other functions in matrixStats
library(matrixStats)
rowMaxs(mat)
rowMaxs(mat, useNames = T)
colMedians(mat)
```

## Iterating over multiple vectors

If you have two or more vectors containing values that you want to iterate over element-wise and process with a function this can be accomplished with `mapply()`.

```r
mapply(FUN, ..., MoreArgs = NULL, SIMPLIFY = TRUE)
```

FUN = function   
... = vectors to iterate over, supplied as named arguments  
MoreArgs = additional arguments with are not vectors supplied as a list   
SIMPLIFY = whether to simplify the result  
       
```{r, fig.width = 10, fig.height = 8}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
shapes <- c(1, 2, 3, 4)

expr_plots <- mapply(plot_expr, # function
                     sys_name = genes, # vector 1 
                     shape = shapes, # vector 2
                     MoreArgs = list(input = brauer_gene_exp), # fixed arguments
                     SIMPLIFY = FALSE) # return a list, rather than coerce to simpler type
wrap_plots(expr_plots)
```

<br>

*Use `mapply()` to create plots for 4 different genes, each with a different point size* 

<details>
  <summary>Show answer</summary>

```{r, fig.width = 10, fig.height = 8}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
sizes <- c(1, 2, 4, 6)

expr_plots <- mapply(plot_expr, # function
                     sys_name = genes, 
                     size = sizes,
                     MoreArgs = list(input = brauer_gene_exp),
                     SIMPLIFY = FALSE)

wrap_plots(expr_plots)
```

</details>

<br>

Alternatively you may find it more readable to use a for loop and a list to store all of the arguments. 


```{r, fig.width = 10, fig.height = 8}
n_plots <- 4
plot_args <- list(
  sys_name = c("YDL104C", "YOR069W", "YLR115W", "YPR036W"),
  size     = c(2, 4, 6, 8),
  shape    = c(1, 2, 3, 4)
)

# preallocate a list of length 4
expr_plots <- vector(mode = "list", length = n_plots)

for(i in seq_len(n_plots)){
  p <- plot_expr(input = brauer_gene_exp,
            sys_name = plot_args$sys_name[i],
            size = plot_args$size[i],
            shape = plot_args$shape[i])
  expr_plots[[i]] <- p
}

wrap_plots(expr_plots)
```


## Using the `map()` family of functions from the `purrr` package

The `purrr` package from the tidyverse provides functions similar to `lapply`, which require even less code than `lapply`.


There is a function for each type of output:

- `map()` makes a list (similar to `lapply()`)  
- `map2()` iterate over two vector/lists by element (similar to `mapply() with 2 vectors`)  
- `pmap()` iterate over each element of a list element-wise (similar to `mapply()`)  

- `map_dfr()` iterate over a vector/list, then combine output into a data.frame using `bind_rows()`  
- `map_cfr()` iterate over a vector/list, then combine output into a data.frame using `bind_cols()`  

- `map_lgl()` iterate over a vector/list, return a logical vector (similar to vapply)  
- `map_int()` iterate over a vector/list, return an integer vector  
- `map_dbl()` iterate over a vector/list, return a double vector  
- `map_chr()` iterate over a vector/list, return a character vector  

Each `map()` function requires two inputs: `map(.x, .f, ...)`  

- `.x` is a list or atomic vector  
- `.f` is a function or formula  

```{r, eval = FALSE}
# We previously lapply to create vectors with different values for mean
vals <- c(1, 50, 20, 5, 70)                  # input
out <- lapply(vals, function(x) rnorm(n = 10, x))

# Using map() we can further simplify this code
# .x indicates where each element of the vector should be inserted
# the ~ is shorthand for an anonymous function e.e. function(.x)
out <- map(.x = vals, .f = ~ rnorm(n = 10, mean = .x))
out <- map(vals, ~rnorm(n = 10, mean = .x))

# You can use brackets to include a multi-line code block
out <- map(vals, ~ {
  rnorm(n = 10, mean = .x)
})

# or use syntax similar to lapply
out <- map(vals, function(input) mean(n = 10, input))

# map() allows for very readable code
# Each element of the vector is passed to the first available argument
out <- map(vals, rnorm, n = 10)
out <- lapply(vals, rnorm, n = 10)
```


## Additional resources related to Programming in R

[Control Structures, from R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/control-structures.html )

[Programming Basics: Introduction to Data Science](http://rafalab.dfci.harvard.edu/dsbook/programming-basics.html#programming-basics)

[Control Flow: Advanced R](https://adv-r.hadley.nz/control-flow.html#control-flow)

<details><summary>Show session info</summary>

```{r code}
sessionInfo()
```
</details>
