---
title: "Class 2: Data wrangling with the tidyverse"
author:
  - name: "Kent Riemondy"
    url: https://github.com/kriemo
    affiliation: "RNA Bioscience Initiative"
    affiliation_url: https://medschool.cuanschutz.edu/rbi
    orcid_id: 0000-0003-0750-1273
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  R.options = list(width = 80)
)
```

*The Rmarkdown for this class is [on github]( https://github.com/rnabioco/bmsc-7810-pbda/blob/main/_posts/2022-10-04-class-2/class-2.Rmd)*

## Introduction to the tidyverse

The [tidyverse](https://www.tidyverse.org/) is a collection of packages that share similar design philosophy, syntax, and data structures. The packages are largely developed by the same team that builds Rstudio. 

Some key packages that we will touch on in this course:

`ggplot2`: plotting based on the "grammar of graphics"  
`dplyr`: functions to manipulate tabular data  
`tidyr`: functions to help reshape data into a tidy format  
`readr`: functions for data import and export  
`stringr`: functions for working with strings  
`tibble`: a redesigned data.frame  


## loading R packages

To use an R package in an analysis we need to load the package using the `library()` function. This needs to be done once in each R session and it is a good idea to do this at the beginning of your Rmarkdown. For teaching purposes I will however sometimes load a package when I introduce a function from a package. 

```{r}
library(readr)
library(dplyr)
library(tibble)
```

## tibble versus data.frame

A `tibble` is a reimagining of the base R `data.frame`. It has a few differences from the `data.frame`.The biggest differences are that it doesn't have `row.names` and it has an enhanced `print` method. If interested in learning more, see the tibble [vignette](https://tibble.tidyverse.org/articles/tibble.html).

Compare `data` to `data_tbl`.

**Note, by default Rstudio displays data.frames in a tibble-like format**

```{r, eval = FALSE}
data <- data.frame(a = 1:3, 
                   b = letters[1:3], 
                   c = Sys.Date() - 1:3, 
                   row.names = c("a", "b", "c"))
data_tbl <- as_tibble(data)
data_tbl
```

When you work with tidyverse functions it is a good practice to convert data.frames to tibbles.

## Convertly a typical data.frame to a tibble

If a data.frame has rownames, you can preserve these by moving them into a column before converting to a tibble using the `rownames_to_column()` from `tibble`.  

```{r}
mtcars # built in dataset, a data.frame with information about vehicles
```

```{r}
mtcars_tbl <- rownames_to_column(mtcars, "vehicle")
mtcars_tbl <- as_tibble(mtcars_tbl)
mtcars_tbl
```

If you don't need the rownames, then you can use the `as_tibble()` function directly. 

```{r}
mtcars_tbl <- as_tibble(mtcars)
```

### Exploring data 

`View()` can be used to open an excel like view of a data.frame. This is a good way to quickly look at the data. `glimpse()` or `str()` give an additional view of the data. 

```r
View(mtcars)
glimpse(mtcars)
str(mtcars)
```

Additional R functions to help with exploring data.frames (and tibbles):

```{r, eval = FALSE}
dim(mtcars) # of rows and columns
nrow(mtcars)
ncol(mtcars)

head(mtcars) # first 6 lines
head(mtcars, n = 2)
tail(mtcars) # last 6 lines
colnames(mtcars) # column names
rownames(mtcars) # row names (not present in tibble)
```

Useful base R functions for exploring values

```{r, eval = FALSE}
mtcars$gear # extract gear column data as a vector
mtcars[, "gear"] # extract gear column data as a vector
mtcars[["gear"]] # extract gear column data as a vector

summary(mtcars$gear) # get summary stats on column

unique(mtcars$cyl) # find unique values in column cyl
length(mtcars$cyl) # length of values in a vector
  
table(mtcars$cyl) # get frequency of each value in column cyl
table(mtcars$gear, mtcars$cyl) # get frequency of each combination of values
```

## Data import using readr

The [`readr`](https://readr.tidyverse.org/) package provides a series of functions for importing or writing data in common text formats.

`read_csv()`: comma-separated values (CSV) files  
`read_tsv()`: tab-separated values (TSV) files  
`read_delim()`: delimited files (CSV and TSV are important special cases)  
`read_fwf()`: fixed-width files  
`read_table()`: whitespace-separated files  

These functions are faster and have better defaults than the base R equivalents (e.g. `read.table`). These functions also directly output tibbles compatible with the tidyverse. 

The [readr checksheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf) provides a concise overview of the functionality in the package. 

To illustrate how to use readr we will load a `.csv` file containing information about flights from 2014. 

First we will download the data. You can download this data manually from [github](https://github.com/arunsrinivasan/flights). Instead we will use R to download the dataset using the `download.file()` base R function.

```{r}
if(!file.exists("flights14.csv")) {
  url <- "https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv" 
  download.file(url, "flights14.csv")
}  
```

You should now have a file called "flights14.csv" in your working directory (the same directory as the Rmarkdown). To read this data into R, we can use the `read_csv()` function. The defaults for this function often work for many datasets. 

```{r}
flights <- read_csv("flights14.csv")
flights
```

There are a few commonly used arguments:
`col_names`: if the data doesn't have column names, you can provide them (or skip them).   

`col_types`: set this if the data type of a column is incorrectly inferred by readr  

`comment`: if there are comment lines in the file, such as a header line prefixed with `#`, you want to skip, set this to `#`.   

`skip`: # of lines to skip before reading in the data.  

`n_max`: maximum number of lines to read, useful for testing reading in large datasets.  


The readr functions will also automatically uncompress gzipped or zipped datasets, and additionally can read data directly from a URL. 

```r
read_csv("https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv")
```

There are equivalent functions for writing data from R to files:
`write_csv`, `write_tsv`, `write_delim`.

## Data import/export for excel files

The `readxl` package can read data from excel files and is included in the tidyverse. The `read_excel()` function is the main function for reading data. 
The `openxlsx` package, which is not part of tidyverse but is on [CRAN](https://ycphs.github.io/openxlsx/index.html), can write excel files. The `write.xlsx()` function is the main function for writing data to excel spreadsheets.

## Data import/export of R objects

Often it is useful to store R objects on disk. These could be large processed datasets, intermediate results, or complex data structures that are not easily stored in rectangular text formats. 

R provides the `readRDS()` and `saveRDS()` functions for storing data in binary formats. 

```{r}
saveRDS(flights, "flights.rds") # save single object into a file
df <- readRDS("flights.rds") # read object back into R
df
```

If you want to save/load multiple objects you can use `save()` and `load()`. 

```{r}
save(flights, df, file = "robjs.rda")  # save flight_df and df
```

`load()` will load the data into the environment with the same objects names used when saving the objects.

```{r}
rm(flights, df)
load("robjs.rda")
```

## Grammar for data manipulation: dplyr

`dplyr` provides a suite of functions for manipulating data 
in tibbles. 

*Rows:    
  - `filter()` chooses rows based on column values  
  - `slice()` chooses rows based on location  
  - `arrange()` changes the order of the rows  
  - `distinct()` selects distinct/unique rows  
  
*Columns:  
  - `select()` changes whether or not a column is included  
  - `rename()` changes the name of columns    
  - `mutate()` changes the values of columns and creates new columns  
  
Groups of rows:  
  - `summarise()` collapses a group into a single row  


## Chaining operations

The `magrittr` package provides the pipe operator `%>%`. This operator allows you to pass data from one function to another. The pipe takes data from the left-hand operation and passes it to the first argument of the right-hand operation. `x %>% f(y)` is equivalent to `f(x, y)`. There is now also a pipe operator in base R (`|>`) which is starting to become more widely used.

The pipe allows complex operations to be conducted without having many intermediate variables. Chaining multiple dplyr commands is a very power and readable 

```{r}
nrow(flights)
flights %>% nrow() # get number of rows
flights %>% nrow(x = .) # the `.` is a placeholder for the data moving through the pipe and is implied
flights %>% colnames() %>% sort() # sort the column names

# you still need to assign the output if you want to use it later
number_of_rows <- flights %>% nrow() 
number_of_rows 
```

### Filter rows

Returning to our `flights` data. Let's use `filter()` to select certain rows. 

`filter(tibble, conditional_expression, ...)`


```{r}
filter(flights, dest == "LAX") #select rows where the `dest` column is equal to `LAX
```

```{r, eval = FALSE}
filter(flights, arr_delay > 200) # flights with arr_delay > 200
filter(flights, distance < 100) # flights less than 100 miles
filter(flights, year != 2014) # if no rows satisfy condition, then an empty tibble
```

Multiple conditions can be used to select rows. For example we can select rows where the `dest` column is equal to `LAX` and the `origin` is equal to `EWR`. You can either use the `&` operator, or supply multiple arguments. 

```{r, eval = FALSE}
filter(flights, dest == "LAX", origin == "EWR")
filter(flights, dest == "LAX" & origin == "EWR")
```

We can select rows where the `dest` column is equal to `LAX` **or** the `origin` is equal to `EWR` using the `|` operator.

```{r, eval = FALSE}
filter(flights, dest == "LAX" | origin == "EWR")
```

The `%in%` operator is useful for identifying rows with entries matching those in a vector of possibilities.

```{r, eval = FALSE}
filter(flights, dest %in% c("LAX", "SLC", "SFO"))
filter(flights, !dest %in% c("LAX", "SLC", "SFO")) # ! will negate
```

Try it out:

- Use filter to find flights to DEN with a delayed departure (`dep_delay`).

```{r}
filter(flights, dest == "DEN", dep_delay > 0)
```

### arrange rows 

`arrange()` can be used to sort the data based on values in a single or multiple columns

`arrange(tibble, <columns_to_sort_by>)`  


For example, let's find the  flight with the shortest amount of air time by arranging the table based on the `air_time` (flight time in minutes).  


```{r}
arrange(flights, air_time) 
```

```{r, eval = FALSE}
arrange(flights, air_time, distance) # sort first on distance, then on air_time

 # to sort in decreasing order, wrap the column name in `desc()`.
arrange(flights, desc(air_time), distance)
```

Try it out:

- Use arrange to rank the data by flight distance (`distance`), rank in ascending order. What flight has the shortest distance?

```{r}
arrange(flights, distance) %>% slice(1) 
```

## Column operations

### select columns

`select()` is a simple function that subsets the tibble to keep certain columns. 

`select(tibble, <columns_to_keep>)`

```{r}
select(flights, origin, dest)
```
the `:` operator can select a range of columns, such as the columns from `air_time` to `hour`. The `!` operator selects columns not listed. 

```{r, eval = FALSE}
select(flights, air_time:hour)
select(flights, !(air_time:hour))
```

There is a  suite of utilities in the tidyverse to help with select columns based on conditions: `matches()`, `starts_with()`, `ends_with()`, `contains()`, `any_of()`, and `all_of()`. `everything()` is also useful as a placeholder for all columns not explicitly listed. See help ?select

```{r, eval = FALSE}
# keep columns that have "delay" in the name
select(flights, contains("delay"))

# select all columns except carrier
select(flights, -carrier)

# reorder columns so that distance and hour are first columns
select(flights, starts_with("di"), ends_with("ay"))
```

## When to quote or not quote?

In general, when working with the tidyverse, you don't need to quote the names of columns. In the example above, we needed quotes because "delay" is not a column name in the flights tibble.  

## Adding new columns with mutate

`mutate()` allows you to add new columns to the tibble. 

`mutate(tibble, new_column_name = expression, ...)`

```{r}
mutate(flights, total_delay = dep_delay + arr_delay)
```

We can't see the new column, so we add a select command to examine the columns of interest.
  
```{r}
mutate(flights, total_delay = dep_delay + arr_delay) %>% 
  select(dep_delay, arr_delay, total_delay)
```

Multiple new columns can be made, and you can refer to columns made in preceding statements. 

```{r, eval = FALSE}
mutate(flights, 
       total_delay = dep_delay + arr_delay,
       rank_delay = rank(total_delay)) %>% 
  select(total_delay, rank_delay)
```

Try it out:

- Calculate the flight time (`air_time`) in hours rather than in minutes, add as a new column.

```{r}
mutate(flights, flight_time = air_time / 60)
```


## Summarizing columns

`summarize()` is a function that will collapse the data from a column into a summary value based on a function that takes a vector and returns a single value (e.g. mean(), sum(), median()). It is not very useful yet, but will be very powerful when we discuss grouped operations. 

```{r}
summarize(flights, 
          avg_arr_delay = mean(arr_delay),
          med_air_time = median(air_time))
```

## Grouped operations

All of the functionality described above can be easily expressed in base R syntax (see examples [here](https://dplyr.tidyverse.org/articles/base.html)). However, where dplyr really shines is the ability to apply the functions above to groups of data within each data frame. 

We can establish groups within the data using `group_by()`. The functions `mutate()`, `summarize()`, and optionally `arrange()` will instead operate on each group independently rather than all of the rows.

Common approaches:
group_by -> summarize: calculate summaries per group
group_by -> mutate: calculate summaries per group and add as new column to original tibble

`group_by(tibble, <columns_to_establish_groups>)`

```{r, eval = FALSE}
group_by(flights, carrier) # notice the new "Groups:" metadata. 

# calculate average dep_delay per carrier
group_by(flights, carrier) %>% 
  summarize(avg_dep_delay = mean(dep_delay)) 

# calculate average arr_delay per carrier at each airport
group_by(flights, carrier, origin) %>% 
  summarize(avg_dep_delay = mean(dep_delay)) 

# calculate # of flights between each origin and destination city, per carrier, and average air time.
 # n() is a special function that returns the # of rows per group
group_by(flights, carrier, origin, dest) %>%
  summarize(n_flights = n(),
            mean_air_time = mean(air_time))  
```

Here are some questions that we can answer using grouped operations in a few lines of dplyr code. Use pipes. 

- What is the average flight `air_time` between each origin airport and destination airport?

```{r}
group_by(flights, origin, dest) %>% 
  summarize(avg_air_time = mean(air_time))
```

- What are the fastest and longest cities to fly between on average? 

```{r}
group_by(flights, origin, dest) %>% 
  summarize(avg_air_time = mean(air_time)) %>% 
  arrange(avg_air_time) %>% 
  head(1)

group_by(flights, origin, dest) %>% 
  summarize(avg_air_time = mean(air_time)) %>% 
  arrange(desc(avg_air_time)) %>% 
  head(1)
```

Try it out:

- Which carrier has the fastest flight (`air_time`) on average from JFK to LAX?

```{r, echo = FALSE}
filter(flights, origin == "JFK", dest == "LAX") %>% 
  group_by(carrier) %>% 
  summarize(flight_time = mean(air_time)) %>% 
  arrange(flight_time) %>% 
  head()
```

- Which month has the longest departure delays on average when flying from JFK to HNL?

```{r, echo = FALSE}
filter(flights, origin == "JFK", dest == "HNL")  %>% 
  group_by(month) %>% 
  summarize(mean_dep_delay = mean(dep_delay)) %>% 
  arrange(desc(mean_dep_delay))
```




## String manipulation

`stringr` is a package for working with strings (i.e. character vectors). It provides a consistent syntax for string manipulation and can perform many routine tasks:


`str_c`: concatenate strings (similar to `paste()` in base R)  
`str_count`: count occurrence of a substring in a string  
`str_subset`: keep strings with a substring  
`str_replace`: replace a string with another string  
`str_split`: split a string into multiple pieces based on a string  


```{r, eval = FALSE}
library(stringr)
some_words <- c("a sentence", "with a ", "needle in a", "haystack")
str_detect(some_words, "needle") # use with dplyr::filter
str_subset(some_words, "needle")

str_replace(some_words, "needle", "pumpkin")
str_replace_all(some_words, "a", "A")

str_c(some_words, collapse = " ")

str_c(some_words, " words words words", " anisfhlsdihg")

str_count(some_words, "a")
str_split(some_words, " ")
```

stringr uses [regular expressions](https://en.wikipedia.org/wiki/Regular_expression) to pattern match strings. This means that you can perform complex matching to the strings of interest. Additionally this means that there are special characters with behaviors that may be surprising if you are unaware of regular expressions.

A useful resource when using regular expressions is https://regex101.com 

```{r, eval = FALSE}
complex_strings <- c("10101-howdy", "34-world", "howdy-1010", "world-.")
# keep words with a series of #s followed by a dash, + indicates one or more occurrences.
str_subset(complex_strings, "[0-9]+-") 

# keep words with a dash followed by a series of #s
str_subset(complex_strings, "-[0-9]+") 

str_subset(complex_strings, "^howdy") # keep words starting with howdy
str_subset(complex_strings, "howdy$") # keep words ending with howdy
str_subset(complex_strings, ".") # . signifies any character
str_subset(complex_strings, "\\.") # need to use backticks to match literal special character
```

Let's use dplyr and stringr together. 

Which destinations contain an "LL" in their 3 letter code?

```{r}
library(stringr)
filter(flights, str_detect(dest, "LL")) %>% 
  select(dest) %>% 
  unique()
```

Which 3-letter destination codes start with H?

```{r}
filter(flights, str_detect(dest, "^H")) %>% 
  select(dest) %>% 
  unique()
```

Let's make a new column that combines the `origin` and `dest` columns. 

```{r}
mutate(flights, new_col = str_c(origin, ":", dest)) %>% 
  select(new_col, everything())
```


<details><summary>Show session info</summary>

```{r code}
sessionInfo()
```
</details>

## Acknowledgements and additional references {.appendix}

The content of this class borrows heavily from previous tutorials:

R code style guide: 
http://adv-r.had.co.nz/Style.html 

Tutorial organization:
https://github.com/sjaganna/molb7910-2019 

Other R tutorials:
https://github.com/matloff/fasteR 
https://r4ds.had.co.nz/index.html 
https://bookdown.org/rdpeng/rprogdatascience/
