---
title: "Class 12: Programming in R (part 2)"
author:
  - name: Ryan Sheridan and Kent Riemondy
date: 2023-12-14
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "left"
)
```

*The Rmarkdown for this class is [on github]( https://github.com/rnabioco/bmsc-7810-pbda/blob/main/_posts/2022-12-05-class-11-programming-in-r-pt2/programming-in-r-pt2.Rmd)*


```{r}
# conditionally download all of the files used in rmarkdown from github
source("https://raw.githubusercontent.com/rnabioco/bmsc-7810-pbda/main/_posts/2023-12-13-class-11-programming-pt-1/download-files.R")
```


<br>

## Introduction

As discussed in the previous class, you should try to limit duplication in your code. One way to do this is by writing functions, another way is through iteration. Reducing code duplication has several benefits:

* Your code easier to read
* You reduce the chance of mistakes from repeated copying and pasting
* It is easier to adapt your code for different requirements

```{r}
library(tidyverse)
library(patchwork)
```

<br>

<br>


## `for` loops

`for` loops allow you to run the same code block repeatedly without copying and pasting. 

```{r}
for (i in 1:4) {
  message("printing ", i, " is fun")
}

values <- c("A", "B", "C", "D")

for(val in values){
  message(val)
}
```

<br>

When iterating over a vector, usually it is most useful to iterate over the index of each element (aka the position in the vector), rather than the values themselves. 

```{r}
for (i in 1:length(values)) {
  val <- values[i]
  message("index = ", i, " value = ", val)
}
```

<br>

We will generally want to store the output generated in the `for` loop. A common paradigm is
to preallocate a place to store the output. This is a faster approach than growing the output at each iteration (for more detail see [this R-bloggers post](https://www.r-bloggers.com/2018/08/growing-objects-and-loop-memory-pre-allocation/)). 

We can generate vectors (and lists) of a given type and length using the `vector()` function. 

```{r}
n <- length(values)

# Make an empty vector that is the same length as values (4)
output <- vector(mode = "character", length = n)

output
for(i in 1:n){ 
  
  # get value at position i
  val <- values[i]
  
  # assign value to output character vector at position i
  output[i] <- tolower(val)
}

output
```

```{r}
# It's helpful to think about what happens during each cycle of the loop
output[1] <- tolower("A")  # i = 1
output[2] <- tolower("B")  # i = 2
output[3] <- tolower("C")  # i = 3
output[4] <- tolower("D")  # i = 4
```

<br> 

Lets use `rnorm()` to create a list of 5 vectors with different values for 'mean'

```{r}
# One way to do this is by copying and pasting
vec_in <- c(1, 50, 20, 5, 70)                # input

out <- vector("list", length(vec_in))        # output

out[[1]] <- rnorm(n = 10, mean = vec_in[1])    
out[[2]] <- rnorm(n = 10, mean = vec_in[2])
out[[3]] <- rnorm(n = 10, mean = vec_in[3])
out[[4]] <- rnorm(n = 10, mean = vec_in[4])
out[[5]] <- rnorm(n = 10, mean = vec_in[5])

out
```

```{r}
# Use a for loop to reduce code duplication
vec_in <- c(1, 50, 20, 5, 70)                  # input

out <- vector("list", length(vec_in))          # output

for (i in 1:length(vec_in)) {                  # sequence
  
  out[[i]] <- rnorm(n = 10, mean = vec_in[i])  # body
  
}
```

<br>

*Write a `for` loop that uses `rnorm()` to create 3 vectors of different lengths. Store the vectors in a list. Use mean = 0 and sd = 1 (the default). *  

<details>
<summary>Show answer</summary>
  
```{r}
vec_in <- c(5, 10, 2)                  # input

n   <- length(vec_in)
out <- vector("list", n)               # output

for (i in 1:length(vec_in)) {          # sequence
  
  out[[i]] <- rnorm(n = vec_in[i])
  
}

out
```

</details>

<br>

So far we have used `1:length(x)` to specify the sequence to iterate over. A better alternative is using `seq_along(x)` instead of `1:length(x)`. This guards against errors when an empty vector is passed to `1:length(x)`.

```{r}
# seq_along() mimics 1:length() for non-empty vectors
vec_in <- c(5, 10, 2)

1:length(vec_in)
seq_along(vec_in)
# seq_along() correctly handles empty vectors
emp_vec <- vector("numeric", 0)

1:length(emp_vec)
seq_along(emp_vec)
```

<br>

<br>

## Using the Brauer data

Using the Brauer gene expression data lets create a figure showing the growth rate vs expression for four genes

```{r}
# Load data
brauer_gene_exp <- read_csv("data/brauer_gene_exp.csv.gz")
```

```{r}
# This is the function we wrote last class
plot_expr <- function(input, sys_name, ...) {
  
  gg_data <- input |>
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  gg_data |>
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(...) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme(legend.position = "none")
}
```

<br>

Lets try this with the copy-and-paste method, storing the plots in a list.

```{r, fig.width = 10, fig.height = 8}
vec_in <- c("YDL104C", "YLR115W", "YMR183C", "YML017W")       # input

out <- vector("list", length(vec_in))                         # output 

out[[1]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[1])  
out[[2]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[2])
out[[3]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[3])
out[[4]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[4])

wrap_plots(out)
```

<br>

*Re-write the code from above using a `for` loop to generate our figure*  

<details>
<summary>Show answer</summary>
  
```{r, fig.width = 10, fig.height = 8}
vec_in <- c("YDL104C", "YLR115W", "YMR183C", "YML017W")  # input

out <- vector("list", length(vec_in))                    # output

for (i in seq_along(vec_in)) {

  out[[i]] <- brauer_gene_exp |>
    plot_expr(sys_name = vec_in[i])

}

wrap_plots(out)
```

</details>

<br>

<br>

## A note on vectorization

*In general you should try to use a vectorized function/approach before using iteration.* Vectorized approaches will be faster and require less code to run. If you are working with a vector or matrix, then there is likely a vectorized operation that can be used. 

There are however a few common places that iteration is used:

  - To process multiple datasets/data.frames. (e.g. apply a function to a list of data.frames)     
  - plotting (e.g. make many plots with a varying input parameter or aesthetic)  
  - perform a custom operation that does not have a vectorized approach 

<br>

<br>

## Introduction to `map()`

`for` loops are a powerful tool to reduce code duplication, however your code can be further simplified using the tidyverse `map()` functions provided in the purrr package. These `map()` functions essentially run `for (i in seq_along(x))` behind the scenes so you don't have to explicitly type this.

There is a function for each type of output:

- `map()` returns a list
- `map_lgl()` returns a logical vector
- `map_int()` returns an integer vector
- `map_dbl()` returns a double vector
- `map_chr()` returns a character vector

Each `map()` function requires two inputs: `map(.x, .f, ...)`

- `.x` is a list or atomic vector
- `.f` is a function or formula

```{r}
# We previously used a for loop to create vectors with different values for mean
vals <- c(1, 50, 20, 5, 70)                  # input

out <- vector("list", length(vals))          # output

for (i in seq_along(vals)) {                 # sequence
  
  out[[i]] <- rnorm(n = 10, mean = vals[i])  # body
  
}

# Using map() we can further simplify this code
# .x indicates where each element of the vector should be inserted
out <- map(
  .x = vals,
  .f = ~ rnorm(n = 10, mean = .x)
)

# You can use brackets to include a multi-line code block
out <- map(vals, ~ {
  
  rnorm(n = 10, mean = .x)

})

# map() allows for very readable code
# Each element of the vector is passed to the first available argument
out <- map(vals, rnorm, n = 10)
```

<br>

*Use `rnorm()` and `map()` to create 3 vectors of different lengths*

<details>
<summary>Show answer</summary>
  
```{r}
lens <- c(5, 10, 2)

out <- map(lens, ~ rnorm(n = .x))

out <- map(lens, rnorm)

out
```

</details>

<br>

*Re-write the code from above using `map()` to generate our growth rate figure*

<details>
<summary>Show answer</summary>

```{r, fig.width = 10, fig.height = 8}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")

expr_plots <- map(
  .x    = genes, 
  .f    = plot_expr, 
  input = brauer_gene_exp
)

wrap_plots(expr_plots)
```

</details>

<br>

Note that we can iterate over lists in addition to vectors. A common operation might be to read in multiple files and perform some operation.

```{r, eval = FALSE}
# get paths to files in "data" directory (dir() is an alias for list.files())
file_names <- dir("data", full.names = TRUE, pattern = ".csv.gz")

# read each file into R and store in a list
lst_of_dfs <- map(file_names, read_csv)

# get nrow of each file
map_int(lst_of_dfs, nrow)

# select 5 random rows
map(lst_of_dfs, slice_sample, n = 5)

# check if any NAs are present
map(lst_of_dfs, ~ !all(complete.cases(.x)))
```

<br>

Note that a `data.frame` is a list in R, such that each column is one element of a list (e.g. see output of `typeof(mtcars)`). So if we use `map()` on a data.frame it will iterate over each column.

```{r, eval = FALSE}
map(mtcars, mean)
map(mtcars, class)
```

<br>

<br>

## Mapping over multiple arguments

If you have two vectors containing values that you want to pass to a function this can be accomplished with `map2()`.

```{r, fig.width = 10, fig.height = 8}
genes  <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
shapes <- c(1, 2, 3, 4)

expr_plots <- map2(genes, shapes, ~ {
  plot_expr(
    input    = brauer_gene_exp,
    sys_name = .x,
    shape    = .y
  )
})

wrap_plots(expr_plots)
```

<br>

*Use `map2()` to create plots for 4 different genes, each with a different point size*

<details>
<summary>Show answer</summary>

```{r, fig.width = 10, fig.height = 8}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
sizes <- c(1, 2, 4, 6)

expr_plots <- map2(genes, sizes, ~ {
  plot_expr(
    input    = brauer_gene_exp,
    sys_name = .x,
    size     = .y
  )
})

wrap_plots(expr_plots)
```

</details>

<br>

`pmap` can be used to map over any number of arguments.

```{r, fig.width = 10, fig.height = 8}
# Create a list of input vectors
genes  <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
sizes  <- c(1, 2, 4, 6)
shapes <- c(1, 2, 3, 4)

plot_args <- list(genes, sizes, shapes)

# Use an argument list with pmap
expr_plots <- pmap(plot_args, ~ {
  plot_expr(
    input = brauer_gene_exp,
    sys_name = ..1,
    size     = ..2,
    shape    = ..3
  )
})

# A simpler way
plot_args <- list(
  sys_name = c("YDL104C", "YOR069W", "YLR115W", "YPR036W"),
  size     = c(2, 4, 6, 8),
  shape    = c(1, 2, 3, 4)
)

expr_plots <- pmap(
  .l = plot_args, 
  .f = plot_expr, 
  input = brauer_gene_exp
)

wrap_plots(expr_plots)
```

<br>

<br>

## Additional resources

[Control Structures, from R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/control-structures.html )

[Programming Basics: Introduction to Data Science](http://rafalab.dfci.harvard.edu/dsbook/programming-basics.html#programming-basics)

[Control Flow: Advanced R](https://adv-r.hadley.nz/control-flow.html#control-flow)

<br>

<br>

<details>
<summary>Show session info</summary>

```{r code}
sessionInfo()
```

</details>
