---
title: "Prerequisite assignment"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**This assignment is due prior to the start of class, Nov 29th before midnight**  

If you run into issues, please follow the instructions provided in the [R package installation and Rstudio overview](https://rnabioco.github.io/bmsc-7810-pbda/posts/2022-10-03-install-r/), or contact the instructors via email (rbi.fellows@cuanschutz.edu). 

This assignment is a simple test to ensure that your R and Rstudio installation is set up. 

1) Change the title of this Rmarkdown to include your name e.g. (`Kent Riemondy Prerequisite assignment`)

2) Add a R code chunk to the document that loads the `tidyverse` package, using the `library` function.

3) Knit this RMarkdown document to html output.

4) Submit the html document to canvas. 


Keep the following code chunk in your rmarkdown, as it will demonstrate which packages you've installed. 

```{r}
sessionInfo()
```

