---
title: "Problem Set 3"
author: "Insert your name here"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup}
library(tidyverse)
```

### Problem Set

Each problem below is worth **2 points**.

For this problem set, you are allowed to help each other.
However, please list who you've worked with and do not post answers in slack.

There are many ways to get the same result.
Try your best to answer the question by fulfilling.
Different plot outputs that still fulfill the task asked for will still be given full credit.
Some questions require a brief written response.
Use the plot to make your best answer and write in the provided area below the plot.
Some questions may use arguments or functions we haven't introduced in class.
This is a great time to use the help functions of RStudio `?`.

**The problem set is due midnight on Dec 8th**

### Grading rubric

-   Everything is good: 2 points.
-   Runs, but wrong answer: 1.5 points
-   Reasonable attempt: 1 points

### Question 1: - Easy Peasy

Create a histogram chart using ggplot2 and the `mtcars` dataset showing the number of cars with different numbers of cylinders.
Adjust the bin value to match the number of cylinder value choices in the dataset.
Add a title and relabel the x axis with a more descriptive label.

```{r}

```

### Question 2: - Not too Bad

Use ggplot2 and the `mtcars` dataset to create a scatter plot showing the relationship between a car's weight and horsepower.
Annotate the plot by labeling the data point with the car name that represents the car with the highest horsepower.
Place a rectangle annotation around the data points that represent the cars with a weight higher than 5.
Label this rectangle with text that says "extremely heavy".
Add a title and relabel the x and y axis with more descriptive labels.

```{r}

```

### Question 3: - The Mean One

Using the `iris` dataset to plot the points of the the petal length for each species.
Make sure to add some jitter as we have done in class.
Now manually calculate the mean of the petal length column of each species independently using the `mean()` function.
Use this value to manually annotate the y position to represent the mean using a red `point` for each species.
Do not use `stat_summary` for this question.

```{r}

```

### Question 4: - The Straightforward One

As we mentioned in class there was an interesting relationship between the quality of cut and the price of the diamond.
Let us revisit that observation in a different way.
First, use the diamonds dataset from the ggplot2 package. Create a histogram (`geom_histogram`) to visualize the distribution of diamond prices, setting a `binwidth` of 100 for detailed granularity. This will give you an initial overview of the overall price distribution.

Next, to specifically examine how this distribution varies with the quality of the cut, employ `facet_wrap` to create separate histograms for each cut category. This will allow us to compare the price distributions across different cut qualities side-by-side.

Add a dashed `geom_vline` layer to each facet to mark the median price for each cut category. This will provide a quick visual reference for comparing median prices across cuts.

```{r}

```

Are the counts of ideal cut diamonds price skewed towards high or low price?
Answer: 

### Question 5: - The Normal One

The price of the diamonds compared to the cut was not enough to understand the its relationship given that ideal cut should represent the best diamonds.
We suspect that the size of the diamond may be a confounding factor.
We can normalize the price to the carat which represents the diamond size.
Repeat the same plot from Question 4 with the following exceptions: Divide the `price` by the `carat`, set the `binwidth` to 0.01, and overwrite the default scale for the x axis to make it on `log10` scale.

```{r}

```

What was the effect of normalizing the data by carat?
Answer: 

### Question 6: - Vroom Vroom

Here we will use a new dataset called `mpg`.
This one is built in so you should be able to access it and find the help and description with `?mpg`.
Using this dataset you will analyze fuel efficiency of several types of cars more modern than from the `mtcars` dataset.
Using a boxplot, plot the `mpg` dataset showing the type of drive train the car uses relative to the highway miles per gallon.
Color (or fill) the plot based upon the type of car. Place the legend at the bottom of the plot. (Hint: use the `theme()` function using the `legend.position` argument. Don't be afraid to search the help!)

```{r}

```

Which type of drive train would appear to be the most fuel efficient?
Answer: 

Which of the types of cars use this drive train in this example?
Answer:

### Question 7: - Shape Up

Using a `subset` of the `mpg` dataset, create a `scatterplot` to analyze the relationship between engine displacement (displ) and highway miles per gallon (hwy). Create the subset focusing on the following five types of cars: 'compact', 'midsize', 'suv', 'pickup', and 'minivan'. Utilize the `shape` aesthetic to differentiate these car types on the `scatterplot`. Then, add a `geom_smooth()` layer to observe the general trend across these car types, using a linear model method (`lm`). Place the legend at the bottom of the plot as we did in the previous problem.

```{r}

```

### Question 8: - Emissions Scandal

Let us compare the `mpg` data for the two time points within the dataset (1999 and 2008).
Plot a boxplot for the model of car and the highway miles per gallon.
Add a `facet_wrap` layer for the year and set the number of rows to 2.
Make sure the plot is legible. Modify the `r{}` code chunk header or swap the coordinates if needed.
Extra challenge if you can figure out how to use the `theme()` function with the `axis.text.x` argument to make the axis labels more readable.

```{r}

```

Which car had the best median highway fuel efficiency in 1999?
Answer: 

Which car had the best median highway fuel efficiency in 2008?
Answer: 

### Question 9: - Publish it!

It is time to create a publication ready plot.
Go back to questions 6, 7, and 8.
Store each of these plots as R objects p1, p2, and p3, respectively.
Make sure that those plots still display in their respective questions.
Now using the `patchwork` library, stitch the plots together such that p1 and p2 are in the same row on the top and p3 is in a row by itself below.
Make sure to add the plot annotations for the subplot letters and add an appropriate title.
Use the `plot_layout()` function from patchwork and modify the argument `heights` to ensure that no subplot is squished. Also be sure to modify the cell `fig.height` or `fig.width` if necessary.
Make sure that `patchwork` is installed with `install.packages("patchwork")` if needed and that it is loaded into your environment with `library(patchwork)`.

```{r}

```

### Question 10: - It's All About Appearances

Please make the most convoluted, distorted, unreadable, useless, ugliest plot possible.
Use any dataset you wish.
Be creative and utilize any of the plot layers and options at your disposal.
Enjoy getting credit for making a visual disaster.

```{r}

```
