---
title: "Problem Set 1"
author: "Insert your name here"
output: html_document
---


### Problem Set

Each problem below is worth **2 points**.

For this problem set, you are allowed to help each other. However, please list who you've worked with and do not post answers in slack. 

Please answer the following questions. Once complete, knit the Rmarkdown to .html and submit your .html document to canvas. 

**The problem set is due 5pm on Dec 1th**

### Grading rubric

- Everything is good: 2 points.
- Runs, but wrong answer: 1.5 points
- Reasonable attempt: 1 points

### q1: 

Using the `state.area` vector, print the 1st, 10th and 50th value of the vector

```{r}
```

### q2 

What is the total area of the states in the 1st, 10ths and 50th positions in the `state.area` vector ?

```{r}

```

### q3
 
What is the area of the largest state?

```{r}

```

### q4

What is the median size of the 10 largest states by area?

```{r}

```


### q4

What is size of the 3rd smallest state?

```{r}

```

### q5

What is the name of the 3rd smallest state by area?

```{r}

```

### q6

How many states have an area greater than 10,000 and less than 500,000?

```{r}

```

### q7

Read the documentation for the `startsWith` function. Using this function, answer the following:

How many state abbreviations (`state.abb` vector) start with `N`?

```{r}

```


### q8

What is the total area of states whose abbreviation starts with `C`?

```{r}

```

### q9

The state.x77 dataset containing various statistics on each state. Answer the following in 
text or code. Consider reading the documentation for assistance (`?state.x77`).

What data structure (e.g. vector, matrix, array, list, etc.) and what data type(s) (e.g. logical, numeric/double, character) are in the `state.x77` object ? (1 point)

Convert the state.x77 to a data.frame, if needed, and answer the following:

How many states have more than 120 days of frost on average (see the `Frost` column)?

```{r}

```

### q10

What is the average population of the Western states? (consider using the state.region vector to assist)

```{r}

```



