---
title: "Problem Set 2"
author: "Insert your name here"
output: html_document
---

```{r setup}
library(tidyverse)
```

### Problem Set

Each problem below is worth **2 points**.

For this problem set, you are allowed to help each other. However, please list who you've worked with and do not post answers in slack. 

There are many ways to get the same result.
Try your best to answer the question by fulfilling.
Different plot outputs that still fulfill the task asked for will still be given full credit.
Some questions require a brief written response. Use the plot to make your best answer and write in the provided area below the plot.
Some questions may use arguments or functions we haven't introduced in class. This is a great time to use the help functions of RStudio `?`.

**The problem set is due 5pm on Dec 9th**

### Grading rubric

- Everything is good: 2 points.
- Runs, but wrong answer: 1.5 points
- Reasonable attempt: 1 points


### Question 1: - Easy Peasy
Create a histogram chart using ggplot2 and the `mtcars` dataset showing the number of cars with different numbers of cylinders. Adjust the bin value to match the number of cylinder value choices in the dataset. Add a title and relabel the x axis with a more descriptive label.

```{r}

```
### Question 2: - Not too Bad

Use ggplot2 and the `mtcars` dataset to create a scatter plot showing the relationship between a car's weight and horsepower. Annotate the plot by labeling the data point with the car name that represents the car with the highest horsepower. Place a rectangle annotation around the data points that represent the cars with a weight higher than 5. Label this rectangle with text that says "extremely heavy". Add a title and relabel the x and y axis with more descriptive labels.

```{r}

```

### Question 3: - The Mean One
Using the `iris` dataset to plot the points of the the petal length for each species. Make sure to add some jitter as we have done in class. Now manually calculate the mean of the petal length column of each species independently using the `mean()` function. Use this value to manually annotate the y position to represent the mean using a red `point` for each species. Do not use `stat_summary` for this question.

```{r}

```


### Question 4: - The Straightforward One
As we mentioned in class there was an interesting relationship between the quality of cut and the price of the diamond. Let us revisit that observation in a different way. Plot the `diamond` dataset and use a `geom_histogram` to plot the price with a binwidth of 100. Use a facet_wrap to look at the relationship of the price to cut. 

```{r}

```
Are the counts of ideal cut diamonds price skewed towards high or low price?
Answer:


### Question 5: - The Normal One
The price of the diamonds compared to the cut was not enough to understand the its relationship given that ideal cut should represent the best diamonds. We suspected in class that the size of the diamond may be a confounding factor. We can normalize the price to the carat which represents the diamond size. Repeat the same plot from Question 4 with the following exceptions: Divide the price by the carat, set the binwidth to 0.01, and overwrite the default scale for the x axis to make it on log10 scale.

```{r}

```
How would you describe the distribution of the ideal cut data now, are they still skewed?
Answer:


### Question 6: - 
Here we will use a new dataset called `mpg`. This one is built in so you should be able to access it and find the help and description with `?mpg`. Using this dataset you will analyze fuel efficiency of several types of more modern cars than the `mtcars` dataset. Using a boxplot, plot the `mpg` dataset showing the type of drive train the car uses relative to the highway miles per gallon. Color the plot based upon the type of car.

```{r}

```
Which type of drive train would appear to be the most fuel efficient? 
Answer:

Which of the types of cars use this drive train in this example?
Answer:

### Question 7: - Shape Up
For this question use the `mpg` dataset to plot a scatterplot of the engine displacement versus the highway miles per gallon fuel efficiency. Utilize the `shape` aesthetic to visualize the types of car on the scatterplot. Finally add a `geom_smooth()` layer to look at the general trend, but only create a single line for the entire set of points, not for each type of car. Set the `method` attribute of `geom_smooth()` to `lm` to use a linear model (straight line).

```{r}

```
### Question 8: - Emissions Scandal
Let us compare the `mpg` data for the two time points within the dataset. Plot a boxplot for the model of car and the highway miles per gallon. Add a `facet_wrap` layer for the year and set the number of rows to 2. Make sure the plot is legible. Modify the `r{}` code chunk header or swap the coordinates if needed.

```{r}

```
Which car had the best median highway fuel efficiency in 1999?
Answer:

Which car had the best median highway fuel efficiency in 2008?
Answer:

### Question 9: - Publish it!
It is time to create a publication ready plot. Go back to questions 6, 7, and 8. Store each of these plots as R objects p1, p2, and p3, respectively. Make sure that those plots still display in their respective questions. Now using the `patchwork` library, stitch the plots together such that p1 and p2 are in the same row on the top and p3 is in a row by itself below. Make sure to add the plot annotations for the subplot letters and add an appropriate title. Make sure that `patchwork` is installed with `install.packages("patchwork")` if needed and that it is loaded into your environment with `library(patchwork)`.

```{r}

```

### Question 10: - It's All About Appearances
Please make the most convoluted, distorted, unreadable, useless, ugliest plot possible. Use any dataset you wish. Be creative and utilize any of the plot layers and options at your disposal. Enjoy getting credit for making a visual disaster.

```{r}

```


